Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-26 rnni_convexity.tex
pass to MDTree

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ between two given trees $T$ and $R$ is computed by sorting the list representations of $T$ and $R$ as it is done in Bubble Sort.
The general idea is to update the list representing $T$ by considering it from beginning to end and swap two neighboured taxa if they appear in a different order in $R$ than in the current list.
Note that each swap of taxa results in a new tree that is added to the path $p$.
After passing the whole list the last element, which is the uppermost taxon, is at the same position as in $R$.
By repeating this at most $n-1$ times all elements the resulting list represents $R$.

@@

Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $T$ and $R$ need to swap at some point on any path between these trees.
And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path indeed.

\begin{algorithm}[H]
\caption{$\csort$($T,R$)}
\label{alg:csort}
\begin{algorithmic}[1]
\STATE $T':= T$, $p := [T']$
\STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
% \STATE $L_{\hat T} :=$ list representation of $\hat T$, $L_R :=$ list representation of $R$
\FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
\STATE $j = 0$ \COMMENT{Special case: cherry}
\IF {$L_{T'}[0]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
\STATE Update $T'$: swap $L_{T'}[0]$ and $L_{T'}[2]$
\ELSIF {$L_{T'}[1]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
\STATE Update $\hat T$: swap $L_{T'}[1]$ and $L_{T'}[2]$
\ENDIF
\STATE $j = 2$
% \IF {$L_R^{-1}[L_{\hat T}[j]] > L_R^{-1}[L_{\hat T}[j+1]]$}
\WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
\STATE Update $T'$: Swap $L_{T'}[j]$ and $L_{T'}[j+1]$
\STATE $p = p+T', j = j + 1$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
Though simple, this algorithm will be crucial for some proofs in Section~\ref{section:caterpillar_convex}.

@@

$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows:
At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
Though this representation appears similar to the list representation of caterpillar trees, it is ambiguous since we do not specify how to order cherry taxa.
$\mdtree$ constructs an output tree $R$ as follows:
Initially, $R$ only consists of a cherry with taxa $l_1$ and $l_2$ such that the parent of these taxa has rank $n-1$.
The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.

>>>>

\todo{Lena: Think if we want to change the following two paragraphs to when we first pull $x_n$ up, then $x_{n-1}$, etc.}
The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ between two given trees $T$ and $R$ is computed by sorting the list representations of $T$ and $R$ as it is done in Bubble Sort.
The general idea is to update the list representing $T$ by considering it from beginning to end and swap two neighboured taxa if they appear in a different order in $R$ than in the current list.
Note that each swap of taxa results in a new tree that is added to the path $p$.
After passing the whole list the last element, which is the uppermost taxon, is at the same position as in $R$.
By repeating this at most $n-1$ times all elements the resulting list represents $R$.

@@

The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
\todo{Lena: If you decide to change the algorithm, mention here that what we've implemented is different from what is described here.}
Note that any algorithm with this property returns a shortest caterpillar path.

\begin{algorithm}[H]
\caption{$\csort$($T,R$)}
\label{alg:csort}
\begin{algorithmic}[1]
\STATE $T':= T$, $p := [T']$
\STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
\FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
\STATE $j = 0$ \COMMENT{Special case: cherry}
\IF {$L_{T'}[0]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
\STATE Update $T'$: swap $L_{T'}[0]$ and $L_{T'}[2]$
\ELSIF {$L_{T'}[1]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
\STATE Update $\hat T$: swap $L_{T'}[1]$ and $L_{T'}[2]$
\ENDIF
\STATE $j = 2$
\WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
\STATE Update $T'$: Swap $L_{T'}[j]$ and $L_{T'}[j+1]$
\STATE $p = p+T', j = j + 1$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.

It is not obvious that the path between $T$ and $R$ returned by $\csort$ has the least possible length among all $\rnni$ paths (not only caterpillar paths), that is, the length of $d(T, R)$.
This fact will be established in Theorem~\ref{}.

@@

$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows:
At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
Though this representation appears similar to the list representation of caterpillar trees, it is ambiguous since we do not specify how to order cherry taxa.
$\mdtree$ constructs an output tree $R$ as follows:
Initially, $R$ only consists two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
